### 2016-01-07 p2/abstract.tex
Minor updates

With distributed computing and mobile applications,
synchronizing diverging replicas of data structures is a more and more common problem.
We use algebraic methods to reason about filesystem operations, 
and introduce a simplified definition of conflicting updates to filesystem trees.
We also define algorithms for update detection and reconciliation
and present rigorous proofs that they not only work as intended,
but also cannot be improved on.

@@

These results are often intuitive,
but providing exact proofs for them is far from trivial.
They contribute to our understanding of this special type of algebraic model,
and toward building more complete algebras
of filesystems 
and extending algebraic approaches to other data storage protocols.
They also form a theoretical basis for
specifying
and guaranteeing the error-free operation
of applications that implement an algebraic approach to synchronization.

>>>>

With distributed computing and mobile applications,
synchronizing diverging replicas of data structures is a more and more common problem.
We use algebraic methods to reason about filesystem operations, 
and introduce a simplified definition of conflicting updates to filesystems.
We also define algorithms for update detection and reconciliation
and present rigorous proofs that they not only work as intended,
but also cannot be improved on.

@@

These results are often intuitive,
but providing exact proofs for them is far from trivial.
They contribute to our understanding of this special type of algebraic model,
and toward building more complete algebras
of filesystem trees 
and extending algebraic approaches to other data storage protocols.
They also form a theoretical basis for
specifying
and guaranteeing the error-free operation
of applications that implement an algebraic approach to synchronization.

### 2016-01-07 p2/algebra.tex
Minor updates

In this algebra, equivalence ($\equiv$) and extension ($\eqext$) would become relations
between sequences, and logical rules involving them
(e.g. $ A\equiv B \Rightarrow S\cc A\cc T\equiv S\cc B\cc T $) would be re-cast as inference rules.
In addition to these, the \namecrefs{ax_separate_commute} listed in \cref{section_axioms}
would become axioms that we accept as true, and from which other true statements can be derived
using the inference rules.
Such a system can allow us to deduce, for example, whether two sequences are equivalent
or one extends the other
without reference to filesystems.

>>>>

In this algebra, equivalence ($\equiv$) and extension ($\eqext$) would become algebraic relations
between sequences, and logical rules involving them
(e.g. $ A\equiv B \Rightarrow S\cc A\cc T\equiv S\cc B\cc T $) would be re-cast as inference rules.
In addition to these, the \namecrefs{ax_separate_commute} listed in \cref{section_axioms}
would become axioms that we accept as true, and from which other true statements can be derived
using the inference rules.
Such a system can allow us to deduce, for example, whether two sequences are equivalent
or one extends the other
without reference to filesystems.

### 2016-01-07 p2/axioms.tex
Finished conclusion
Minor updates

So that we could describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[Sequences of commands: $\cc$ and $\emptyseq$]
We use ``$\cc$'' to concatenate commands to form a sequence, or concatenate sequences to form a longer sequence:
$S=\cbdaa{n}{\vald}\cc\cbfaa{m}{\valf}$, or $S'=S\cc\cbba{o}$,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
Sequences of commands form a free semigroup, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which by definition leaves all filesystems unchanged.
\end{mydef}

The following concepts 
(which echo the ones defined in \cite{NREC})
allow us to investigate sequences independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences of commands.

@@

\begin{mydef}[$\eqext$]
% Relation in the algebra
$A\eqext B$ means that $B$ extends $A$,
% intended interpretation:
that is, they behave in the same way
on any filesystem $A$ does not break:
$\forall \FS: A\FS\neq\fsbroken\Rightarrow A\aFS=B\aFS$.
It is a reflexive and transitive relation,
and we can see that
% inference rules:
$ A\eqext B \Rightarrow S\cc A\cc T\eqext S\cc B\cc T$
and
$ A\equiv B \Rightarrow A\eqext B$.
\end{mydef}

@@

We methodically investigated of all possibilities using a computer 
program\footnote{The program is accessible on-line at \\
https://github.com/csirmaz/AlgebraicSyncPaper/blob/master/p2/prove.py.}
to determine
which pairs of commands cause errors all the time,
which can be simplified to one or no commands, and which can be reversed
without any change in their overall effect.
Our basis for this investigation was the model of filesystems introduced in this paper.
These properties of command pairs are crucial as they determine
how a set of commands can be re-ordered to be applied to a filesystem
during synchronization, and what commands will never be compatible.
Below we list a number of statements derived using this method.

@@

\begin{myax}\label{ax_same_emptyseq}
Commands on the same node simplify:
$\cxynv\cc\czwnv \eqext \emptyseq$ where $Y=Z$ and $X=W=\ccharb$ 
or $X=W=\cchard$.
\end{myax}

@@

% TODO We will only use these except for...

>>>>

So that we can describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[Sequences of commands: $\cc$ and $\emptyseq$]
We use ``$\cc$'' to concatenate commands to form a sequence, or concatenate sequences to form a longer sequence:
$S=\cbdaa{n}{\vald}\cc\cbfaa{m}{\valf}$, or $S'=S\cc\cbba{o}$,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
Sequences of commands form a free semigroup, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which by definition leaves all filesystems unchanged.
\end{mydef}

The following two relations 
(which echo the ones defined in \cite{NREC})
allow us to investigate sequences independently of filesystems.
In the definitions, $A,B,S,T$ stand for arbitrary sequences of commands.

@@

\begin{mydef}[$\eqext$]
% Relation in the algebra
$A\eqext B$ means that $B$ extends $A$,
% intended interpretation:
that is, they behave in the same way
on any filesystem $A$ does not break:
$\forall \FS: A\FS\neq\fsbroken\Rightarrow A\aFS=B\aFS$.
It is a reflexive and transitive relation,
and we can see that
% inference rules:
$ A\eqext B \Rightarrow S\cc A\cc T\eqext S\cc B\cc T$,
and
$ A\equiv B \Rightarrow A\eqext B$.
\end{mydef}

@@

We methodically investigated of all possibilities using a computer 
program\footnote{The program is accessible on-line at \\
https://github.com/csirmaz/AlgebraicSyncPaper/blob/master/p2/prove.py.}
to determine
which pairs of commands cause errors all the time,
which can be simplified to one or no commands, and which commute and can be reversed
without any change in their overall effect.
Our basis for this investigation was the model of filesystems introduced in this paper.
These properties of command pairs are crucial as they determine
how a set of commands can be re-ordered to be applied to a filesystem
during synchronization, and what commands will never be compatible.
Below we list a number of statements derived using this method.

@@

\begin{myax}\label{ax_same_emptyseq}
Commands on the same node simplify:
$\cxynv\cc\czwnv \eqext \emptyseq$ where $Y=Z$, and $X=W=\ccharb$ 
or $X=W=\cchard$.
\end{myax}

@@

### 2016-01-07 p2/conclusion.tex
Minor updates

Apart from constructing an algebra,
there are many other ways in which further work can extend the current results.
An important extension would be to
consider reconciling not only two, but more replicas in a single step and
prove the correctness and maximality of the algorithm proposed.

@@

An extension of our results could also entail relaxing
the simplifying assumptions we used.
Identifying which results depend on the assumption that $|\setd|=1$
and devising updated proofs and algorithms,
or specifying a pre-processing step that reconciles meta-information in directories
can enable the synchronizer
to identify and resolve a wider range of conflicts.
Similarly, designing pre- and post-processors that
present removing and creating a node
not as separate updates but as a single rename (move),
could contribute to the user-friendliness of the reconciler
and help avoid human error.

>>>>

Apart from constructing an algebra,
there are many other ways in which further work can extend the current results.
An important extension would be to
consider reconciling not only two, but more replicas in a single step and
prove the correctness and maximality of the algorithm proposed,
or show that it is impossible to do so.

@@

An extension of our results could also entail relaxing
the simplifying assumptions we used.
Identifying which results depend on the assumption that $|\setd|=1$
and devising updated proofs and algorithms for when $|\setd|>1$,
or specifying a pre-processing step that reconciles meta-information in directories
can enable the synchronizer
to identify and resolve a wider range of conflicts.
Similarly, designing pre- and post-processors that
present removing and creating a node
not as separate updates but as a single rename (move),
could contribute to the user-friendliness of the reconciler
and help avoid human error.

### 2016-01-07 p2/def-commands.tex
Minor updates

To define the exact effect of the commands we use
the \emph{replacement operator}  which changes the
value of the filesystem at a given node only.

\begin{mydef}[Effect of commands]
We use a \emph{replacement operator} of the form
$\FS[\valv/n]$ to denote a filesystem derived from $\FS$:
\[ \FS[\valv/n](m) =
   \begin{cases}
   \valv &\mbox{if~} m=n\\
   \FS(m) &\mbox{otherwise.}
   \end{cases}
\]
The effect of the commands is then as follows:
\begin{align*}
&\cbrk\aFS = \fsbroken \\
&\cxynv\aFS = 
   \begin{cases}
   \fsbroken &\mbox{if~} \FS=\fsbroken\\
   \fsbroken &\mbox{if~} \FS(n)\not\in\setvx{X}\\
   \fsbroken &\mbox{if~} \FS[\valvy/n] \mbox{~violates the tree property}\\
   \FS[\valvy/n] &\mbox{otherwise.}
   \end{cases}
\end{align*}
\end{mydef}

>>>>

\begin{mydef}[Effect of commands]
We use a \emph{replacement operator} of the form
$\FS[\valv/n]$ to denote a filesystem derived from $\FS$ 
by replacing its value at $n$ with $\valv$:
\[ \FS[\valv/n](m) =
   \begin{cases}
   \valv &\mbox{if~} m=n\\
   \FS(m) &\mbox{otherwise.}
   \end{cases}
\]
The effect of the commands is then as follows:
\begin{align*}
&\cbrk\aFS = \fsbroken \\
&\cxynv\aFS = 
   \begin{cases}
   \fsbroken &\mbox{if~} \FS=\fsbroken\\
   \fsbroken &\mbox{if~} \FS(n)\not\in\setvx{X}\\
   \fsbroken &\mbox{if~} \FS[\valvy/n] \mbox{~violates the tree property}\\
   \FS[\valvy/n] &\mbox{otherwise.}
   \end{cases}
\end{align*}
\end{mydef}

### 2016-01-07 p2/intro.tex
Minor updates

In this paper we investigate a command-based approach
to synchronizing filesystem trees.
Presented with multiple copies (replicas) of the same filesystem that have been independently modified,
we regard the aim of the synchronizer to modify these replicas further
to make them as similar as possible.
The synchronization algorithm we describe
follows the two main steps described by Balasubramaniam and Pierce \cite{BP}:
update detection, where we identify modifications that have been applied to the replicas
since they diverged and represent these as sequences of commands,
and reconciliation, where we identify commands that can be propagated to other replicas and do so.
The remaining commands, which could not be propagated, represent conflicting updates;
resolving these requires a separate system or human intervention.

@@

A central problem of command-based synchronizers 
during both update detection and reconciliation
is the ordering (scheduling) of commands.
If update detection is based on comparing the original state of the filesystem
to its new state, then we can easily collect a set of updates,
but we need to order these in a way that they could be applied
to the filesystems without causing an error 
(for example, a directory needs to be created before a file can be created under it).
Similarly, during reconciliation when we consider all the commands that
have been applied to the different replicas, we need to find a way of
merging these and creating a global ordering that we can apply
to the filesystems.
In fact, an insight in \cite{NREC} is that
if the ordering changes the effect of two commands,
then under certain circumstances they are not compatible and will give rise to conflicts.
Accordingly, in this paper command-pairs and their properties
play a special role
as we aim to find valid orderings on filesystem trees.

In terms of the usual classification of synchronizers \cite{TSR, PV, SSH},
our approach to reconciliation is therefore operation-based as we investigate sets of commands,
while the update detection algorithm makes the full synchronizer 
similar to state-based synchronizers 
as its starting point is the states of the replicas themselves.
Moreover, our ordering algorithms only depend on properties of commands 
such as commutativity or idempotence,
and therefore they can be classified as semantic schedulers.
This is in opposition to syntactic scheduling, which uses 
the origin and original temporal order of updates, 
and which may give rise to otherwise resolvable conflicts \cite{SSH}.
In general, ordering plays a central role in operation-based synchronizers.
We refer to a survey by Saito and Shapiro of so-called optimistic replication algorithms \cite{SSH};
and, in particular,
to IceCube, where multiple orders are tested to find an acceptable one (\cite{KRSD}, see also \cite{MPV}),
or Bayou, where reconciling updates happens by redoing them in a globally determined order \cite{TTPDSH}.

>>>>

In this paper we investigate a command-based approach
to synchronizing filesystem trees.
Presented with multiple copies (replicas) of the same filesystem that have been independently modified,
we regard the aim of the synchronizer to modify these replicas further
to make them as similar as possible.
The synchronization algorithm we describe
follows the two main steps described by Balasubramaniam and Pierce \cite{BP}:
update detection, where we identify modifications that have been applied to the replicas
since they diverged and represent them as sequences of commands;
and reconciliation, where we identify commands that can be propagated to other replicas and do so.
The remaining commands, which could not be propagated, represent conflicting updates;
resolving these requires a separate system or human intervention.

@@

A central problem of command-based synchronizers 
during both update detection and reconciliation
is the ordering (scheduling) of commands.
If update detection is based on comparing the original state of the filesystem
to its new state, then we can easily collect a set of updates,
but we need to order these in a way that they could be applied
to the filesystem without causing an error 
(for example, a directory needs to be created before a file can be created under it).
Similarly, during reconciliation when we consider all the commands that
have been applied to the different replicas, we need to find a way of
merging these and creating a global ordering that we can apply
to the filesystems.
In fact, an insight in \cite{NREC} is that
if the ordering changes the effect of two commands,
then under certain circumstances they are not compatible and will give rise to conflicts.
Accordingly, in this paper the commutativity of command pairs and their other properties
play a special role.

In terms of the usual classification of synchronizers \cite{TSR, PV, SSH},
our approach to reconciliation is therefore operation-based as we investigate sets of commands,
while the update detection algorithm makes the full synchronizer 
similar to state-based synchronizers 
as its starting point is the states of the replicas themselves.
Moreover, our ordering algorithms only depend on properties of commands 
such as commutativity or idempotence,
and therefore they can be classified as semantic schedulers.
This is in opposition to syntactic scheduling, which uses 
the origin and the original temporal order of updates, 
and which may give rise to otherwise resolvable conflicts \cite{SSH}.
In general, ordering plays a central role in operation-based synchronizers.
We refer to an excellent survey by Saito and Shapiro of so-called optimistic replication algorithms \cite{SSH};
and, as examples,
to IceCube, where multiple orders are tested to find an acceptable one (\cite{KRSD}, see also \cite{MPV}),
or Bayou, where reconciling updates happens by redoing them in a globally determined order \cite{TTPDSH}.

### 2016-01-07 p2/ordering.tex
Minor updates

Furthermore,
we prove this proposition for sequences, not sets, as if all sequences must contain a command on $\parent(n)$,
then so must sets because otherwise there would be no order in which they can be
applied to any filesystem.

@@

To order the commands in the component,
select the command with the topmost node, so that there would be no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\caa{X}{\ccharb}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any ordering to break all filesystems.

@@

As a specific case, we can regard a simple sequence of commands $S$
as a set and determine its $\orderset{S}$.
An important property of simple sequences is that
all of their valid orderings are equivalent,
and all equivalent sequences are mere reorderings.
To prove this, first we prove the following:

\begin{mylem}\label{equiv_simple_same_commands}
If two simple sequences $A$ and $B$ are equivalent
and do not break all filesystems ($A\equiv B\nequiv\cbrk$),
then they must contain the same commands.
\end{mylem}
\begin{proof}
We use an inverse proof. Let $A$ and $B$ simple sequences
such that $A\equiv B\nequiv\cbrk$,
and $\FS$ be a filesystem that they do not break.

We also assume that they do not contain the same sequences.
Without loss of generality, we can assume
that $A$
contains $\cxynv$, and $B$ either contains a different command
$\czwnv$ on $n$, or no command on $n$ at all.
As $A$ is simple, we know that $\cxynv$ is not an assertion command,
and therefore $X\neq Y$ and $\FS(n)\neq\valvy$ as it has type $X$.
If so, then if $B$ has no command on $n$, then $B\aFS(n)=\FS(n)$ and
$A$ and $B$ cannot be equivalent.
Otherwise we know that $Z=X$ as $B$ does not break $\FS$ either,
and that $\valvw=\valvy$ and therefore $W=Y$ as $B\aFS(n)$ must
be $\valvy$ and $B$ only has one command on $n$.
However, then $\czwnv=\cxynv$, which is a contradiction.
\end{proof}

@@

If, for example, in a component or sub-component
made up of construction pairs,
the command on the topmost node, $\cxynv$, was not before all other commands
in the component as described by the algorithm,
then any resulting sequence would necessarily contain
a $\czwmv$ where $n\descendant m$, potentially followed by commands
on incomparable nodes, and then followed by $\cxynv$.
From \cref{ax_separate_commute} we know we can convert this sequence
to an equivalent sequence by swapping the commands on incomparable nodes with
$\czwmv$, which sequence would therefore have the commands on
$m$ and $n$ next to each other. Then from \cref{ax_distantrel_breaks,ax_directparent_breaks}
we know the sequence would break all filesystems
as a reversed construction pair is never a valid destruction pair.

We can prove that an order not generated by the algorithm
in case of components made of destruction pairs breaks all filesystems
in a similar way.

>>>>

Furthermore,
we prove this proposition for sequences, not sets, as if all sequences must contain a command on $\parent(n)$,
then so must a set because otherwise there would be no order in which it can be
applied to any filesystem.

@@

To order the commands in the component,
select the command with the topmost node, so that there is no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\caa{X}{\ccharb}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any ordering to break all filesystems.

@@

As a specific case, we can regard a simple sequence of commands $S$
as a set and determine its $\orderset{S}$.
\Cref{simple_reorder_equiv} is
an important property of simple sequences: namely, that
all of their valid orderings are equivalent,
and all equivalent sequences are mere reorderings.
To prove this, first we prove the following:

\begin{mylem}\label{equiv_simple_same_commands}
If two simple sequences $A$ and $B$ are equivalent
and do not break all filesystems ($A\equiv B\nequiv\cbrk$),
then they must contain the same commands.
\end{mylem}
\begin{proof}
We use an inverse proof. Let $A$ and $B$ be simple sequences
such that $A\equiv B\nequiv\cbrk$,
and $\FS$ be a filesystem that they do not break.

We also assume that they do not contain the same commands.
Without loss of generality, we can assume
that $A$
contains $\cxynv$, and $B$ either contains a different command
$\czwnv$ on $n$, or no command on $n$ at all.
As $A$ is simple, we know that $\cxynv$ is not an assertion command,
and therefore $X\neq Y$ and $\FS(n)\neq\valvy$ as it has type $X$.
If so, then if $B$ has no command on $n$, then $B\aFS(n)=\FS(n)$ and
$A$ and $B$ cannot be equivalent.
Otherwise we know that $Z=X$ as $B$ does not break $\FS$ either,
and that $\valvw=\valvy$ and therefore $W=Y$ as $B\aFS(n)$ must
be $\valvy$ and $B$ only has one command on $n$.
However, then $\czwnv=\cxynv$, which is a contradiction.
\end{proof}

@@

If, for example, in a component or sub-component
made up of construction pairs,
the command on the topmost node, $\cxynv$, was not before all other commands
in the component as prescribed by the algorithm,
then any resulting sequence would necessarily contain
a $\czwmv$ where $n\descendant m$, potentially followed by commands
on incomparable nodes, and then followed by $\cxynv$.
From \cref{ax_separate_commute} we know we can convert this sequence
to an equivalent sequence by swapping the commands on incomparable nodes with
$\czwmv$, which sequence would therefore have the commands on
$m$ and $n$ next to each other. Then from \cref{ax_distantrel_breaks,ax_directparent_breaks}
we know the sequence would break all filesystems
as a reversed construction pair is never a valid destruction pair.

We can prove that an order not generated by the algorithm
in the case of a component made of destruction pairs breaks all filesystems
in a similar way.

### 2016-01-07 p2/works-inverse.tex
Minor updates

\begin{mydef}[$\worksc{x}{y}$]
For two sets of sequences, $\worksc{A_1,A_2,\ldots,A_k}{B_1,B_2,\ldots,B_l}$ means that 
all of $A_1,\ldots,A_k$ work where all of $B_1,\ldots,B_l$ work,
that is,
% intended interpretation:
\begin{align*}
\forall \FS:{}& 
B_1\FS\neq\fsbroken \wedge \cdots \wedge B_k\FS\neq\fsbroken\\
&\Rightarrow\\
&A_1\FS\neq\fsbroken \wedge \cdots \wedge A_l\FS\neq\fsbroken
\end{align*}
As above, the order of the sequences in the sets is not relevant.
\end{mydef}

@@

\begin{mylem}\label{worksinputmatch}
If $A$ and $B$ are minimal sequences, $\works{A,B}$,
and there are commands on node $n$ in both $A$ ($\cxynv\in A$) and $B$ ($\czwnv\in B$)
then the input types of these commands must match ($X=Z$).
\end{mylem}
\begin{proof}
This result is similar to \cref{equiv_simple_same_commands} and
is easily shown using an indirect proof: if $X\neq Z$, then there is no filesystem that
either $\cxynv$ or $\czwnv$ would not break, 
and consequently $A$ and $B$ cannot work on the same filesystem.
\end{proof}

>>>>

\begin{mydef}[$\worksc{x}{y}$]
For two sets of sequences, $\worksc{A_1,A_2,\ldots,A_k}{B_1,B_2,\ldots,B_l}$ means that 
all of $A_1,\ldots,A_k$ work where all of $B_1,\ldots,B_l$ work,
that is,
% intended interpretation:
\begin{align*}
\forall \FS:{}& 
B_1\FS\neq\fsbroken \wedge \cdots \wedge B_k\FS\neq\fsbroken\\
&\Rightarrow\\
&A_1\FS\neq\fsbroken \wedge \cdots \wedge A_l\FS\neq\fsbroken.
\end{align*}
As above, the order of the sequences in the sets is not relevant.
\end{mydef}

@@

\begin{mylem}\label{worksinputmatch}
If $A$ and $B$ are minimal sequences, $\works{A,B}$,
and there are commands on node $n$ in both $A$ ($\cxynv\in A$) and $B$ ($\czwnv\in B$),
then the input types of these commands must match ($X=Z$).
\end{mylem}
\begin{proof}
This result is similar to \cref{equiv_simple_same_commands} and
is easily shown using an indirect proof: if $X\neq Z$, then there is no filesystem that
either $\cxynv$ or $\czwnv$ would not break, 
and consequently $A$ and $B$ cannot work on the same filesystem.
\end{proof}
